Adding selection to \stdpopsim: oh how the tables have turned

### 2024-11-11 analysis2_manuscript.tex
first two paras

\documentclass[hidelinks]{article}
\usepackage[letterpaper,margin=1.0in]{geometry}
\usepackage[utf8]{inputenc}
\pagenumbering{arabic}
\usepackage{authblk}
\usepackage{graphicx}
\usepackage[singlelinecheck=false]{caption} % singlelinecheck makes single line caption left aligned instead of centered
\usepackage{subcaption}
\usepackage{amsmath}
\usepackage[round]{natbib}
\usepackage{fancyhdr}
\usepackage{longtable}
\usepackage{booktabs}
% hyperlinks
\usepackage{hyperref}

\usepackage{xspace}
\usepackage{mathrsfs}
\usepackage{graphicx}

\pagestyle{fancy}
\fancyhead[R]{\textbf{Expanding \stdpopsim}}

@@

% bibliography
\usepackage[round]{natbib}   % omit 'round' option if you prefer square brackets
\bibliographystyle{plainnat}

@@

@@

% \author[1,+]{M. Elise Lauterbur}
% \author[2,*]{Maria Izabel A. Cavassim}
% \author[3,*]{Ariella L. Gladstein}
 \author[4,*]{Graham Gower}
 \author[5*]{Nathaniel S. Pope}
% \author[6,*]{Georgia Tsambos}
% \author[5,7]{Jeff Adrion}
% \author[5]{Saurabh Belsare}
% \author[8]{Arjun Biddanda}
% \author[5]{Victoria Caudill}
% \author[9]{Jean Cury}
% \author[10]{Ignacio Echevarria}
% \author[11]{Benjamin C. Haller}
% \author[12,13]{Ahmed R. Hasan}
% \author[14,15]{Xin Huang}
% \author[16]{Leonardo Nicola Martin Iasi}
% \author[17]{Ekaterina Noskova}
% \author[18]{Jana Obšteter}
% \author[19]{Vitor Antonio Corrêa Pavinato}
% \author[20,21]{Alice Pearson}
% \author[22,23]{David Peede}
% \author[24]{Manolo F. Perez}
% \author[5]{Murillo F. Rodrigues}
% \author[5]{Chris C. R. Smith}
% \author[25]{Jeffrey P. Spence}
% \author[5]{Anastasia Teterina}
% \author[5]{Silas Tittes}
% \author[26]{Per Unneberg}
% \author[27]{Juan Manuel Vazquez}
% \author[28]{Ryan K. Waples}
% \author[29]{Anthony Wilder Wohns}
% \author[30]{Yan Wong}
% \author[31]{Franz Baumdicker}
% \author[32]{Reed A. Cartwright}
% \author[33]{Gregor Gorjanc}
% \author[34]{Ryan N. Gutenkunst}
% \author[30]{Jerome Kelleher}
% \author[5]{Andrew D. Kern}
% \author[35]{Aaron P. Ragsdale}
% \author[5,36]{Peter L. Ralph}
% \author[37]{Daniel R. Schrider}
% \author[38,+]{Ilan Gronau}

@@

% \affil[30]{\small{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery, University of Oxford, Oxford OX3 7LF, UK}}

@@

% \affil[35]{\small{Department of Integrative Biology, University of Wisconsin-Madison, Madison WI, USA}}
% \affil[36]{\small{Department of Mathematics, University of Oregon, Eugene OR 97402, USA}}
% \affil[37]{\small{Department of Genetics, University of North Carolina at Chapel Hill, Chapel Hill NC 27599, USA}}
% \affil[38]{\small{Efi Arazi School of Computer Science, Reichman University, Herzliya, Israel}}

@@

\section*{Introduction}
    \label{introduction}
    % natural selection
    Natural selection is a fundamental force in evolution, shaping the
    genetic diversity of populations and driving the adaptation of
    species to their environments. The effects of natural selection
    on genetic variation are complex, and can be difficult to disentangle
    from other evolutionary processes such as mutation, recombination,
    and genetic drift \cite[e.g.,][]{gillespie1991causes}.
    For instance changes in population size can lead to fluctuations
    in genetic diversity across a recombining chromosome 
    that can mimic the effects of selection \citep{simonsen1995properties, barton1998effect},
    and lead to spurious inferences about the strength and targets of genetic adaptation
    \cite{simonsen1995properties,akey2004population,nielsen2005genomic}.
    In turn selection can confound our ability to infer demographic 
    history from allele frequencies \citep{ewing2016consequences} and
    estimates of inverse coalescent rate \citep{schrider2016effects}.
    Thus it is imperative to joingly account for the effects of selection
    and demography when inferring evolutionary history from genetic data
    however this is a challenging task from a modeling perspective.

    % simulation
    To meet this need the field of population genetics
    has turned to simulation.
    Simulation has become a critical tool for interpretation, 
    analysis, and exploration of realistic evolutionary models.
    Simulation in population genetics has a long history 
    including both backward in time coalescent simulations
    \citep{kingman1982genealogy,hudson1983testing, hudson1990gene}
    and forward in time simulations of complex demography and selection
    \citep[e.g.,][]{gillespie1984molecular,thornton2014c++, haller2019slim}.
    The development of simulation tools has been driven by the need to
    understand the effects of complex evolutionary processes on genetic
    variation \citep[e.g.,][]{galloway2020few}, to provide a null model for hypothesis testing
    \citep[e.g.,][]{hudson1992statistical}, to
    to explore the power and limitations of statistical methods \cite[e.g.,]{przeworski2002signature},
    and increasingly to provide a basis for machine learning and other
    simulation-based inference methods \citep[e.g.,]{kern2018diplos}.
    While this is so, joint simulation of complex demography and selection
    is challenging, and requires a deep understanding of the underlying
    evolutionary processes, as well as a number of parameter choices including
    the strength of selection, the distribution of fitness effects, and the
    recombination rate, as well as a parameterized model of demography.
    This is dounting task for many researchers, and can be a barrier to
    the adoption of simulation-based methods in population genetics.
    Further, sharing and comparing simulation results among studies can 
    itself be challenging, as different simulation tools may use different
    models and parameterizations, and may not be directly comparable.
    This is a major limitation for the field, as it can make it difficult
    to assess the robustness of results, and to compare the results of
    different studies.

    % stdpopsim
    A lingering challenge with simulation in population genetics has been
    reproducibility and the ability to share and compare results among 
    researchers. This challenge has been addressed in part by the development
    of community resources for sharing and distributing simulation software
    via the \stdpopsim project \citep{adrion2020community}. \stdpopsim
    provides a standardized interface for accessing a wide range of
    population genetic models, and has begun to be be widely adopted by the community.
    While that is so, the current version of \stdpopsim does not include
    models of selection, which is a major limitation for many applications
    of population genetic simulation. In particular, modeling selection
    through simulation is critical for understanding the processes such
    as adaptation, the effects of selective sweeps, and the impact of
    background selection on genetic diversity. Ideally, one would like
    to have a single, unified framework for simulating both neutral and
    non-neutral evolutionary processes, and to be able to compare the
    results of these simulations to empirical data in a manner that is
    accessible to a wide range of researchers and highly reproducible. 
    Further the framework should include the complex realities of 
    genomes, including heterogenous recombination rates, 
    variation in the density of functional elements, and relevant
    population size histories.

\section*{outline}
    \label{outline}
    \begin{enumerate}
        \item Single-population demographic inference methods
        \begin{enumerate}
            \item mscm2
            \item stairwaiplot
            \item GONE
            \item SMC++
        \end{enumerate}
        \item Multi-population demographic inference methods
        \begin{enumerate}
            \item dadi
            \item fastsimcoal
            \item momi2
        \end{enumerate}
        \item DFE inference methods
        \begin{enumerate}
            \item dadi
            \item polyDFE
            \item DFE-alpha
            \item GRAPES
        \end{enumerate} 
        \item sweeps
        \begin{enumerate}
            \item compare effect of BGS on sweep detection
            \item compare power in different pops
            \item compare methods
        \end{enumerate}
    \end{enumerate}

\section*{Additions to stdpopsim}
    \label{additions}

    \begin{figure}[ht]
        \centering
        \includegraphics[width=\textwidth]{figures/HomSap/estimated_Ne_t_final}
        \caption{
        \label{fig:demography}
        Performance of methods to infer human OOA model
        with and without background selection on exons.
        }
    \end{figure}

    \begin{figure}[ht]
        \centering
        \includegraphics[width=\textwidth]{figures/HomSap/dfe.inference.benchmark}
        \caption{
        \label{fig:dfe_humans}
        Performance of methods to infer distribution of fitness effects (DFE).
        }
    \end{figure}

    \begin{figure}[ht]
        \centering
        \includegraphics[width=\textwidth]{figures/PhoSin/dfe.inference.benchmark}
        \caption{
        \label{fig:dfe_vaquita}
        Performance of methods to infer distribution of fitness effects (DFE).
        }
    \end{figure}

\section*{Conclusion}
    \label{conclusion}

>>>>

\documentclass[hidelinks]{article}
\usepackage[letterpaper,margin=1.0in]{geometry}
\usepackage[utf8]{inputenc}
\pagenumbering{arabic}
\usepackage{authblk}
\usepackage{graphicx}
\usepackage[singlelinecheck=false]{caption} % singlelinecheck makes single line caption left aligned instead of centered
\usepackage{subcaption}
\usepackage{amsmath}
\usepackage{fancyhdr}
\usepackage{longtable}
\usepackage{booktabs}
% hyperlinks
\usepackage{hyperref}
\usepackage{wrapfig}
\usepackage{xspace}
\usepackage{mathrsfs}
\usepackage{graphicx}
\usepackage{lipsum}
\usepackage{makecell}

\pagestyle{fancy}
\fancyhead[R]{\textbf{\stdpopsim with selection}}

@@

% bibliography
\usepackage[
    backend=biber,
    natbib=true,
    style=authoryear,
    ]{biblatex}
\addbibresource{references.bib}

%\usepackage[round]{natbib}   % omit 'round' option if you prefer square brackets
%\bibliographystyle{plainnat}

@@

% add commands for msmc2, stairway plot, gone, and smc++
\newcommand{\msmc}{\texttt{msmc2}\xspace}
\newcommand{\stairway}{\texttt{stairwayplot}\xspace}
\newcommand{\gone}{\texttt{GONE}\xspace}
\newcommand{\smcpp}{\texttt{SMC++}\xspace}

% add commands for the DFE inference methods
\newcommand{\dfe}{\texttt{dfe-alpha}\xspace}
\newcommand{\polydfe}{\texttt{polyDFE}\xspace}
\newcommand{\dadi}{$\partial$\texttt{a}$\partial$\texttt{i}\xspace}
\newcommand{\grapes}{\texttt{GRAPES}\xspace}

% add commands for the sweep detection methods
\newcommand{\sweepfinder}{\texttt{sweepfinder2}\xspace}
\newcommand{\diploshic}{\texttt{diploshic}\xspace}

@@

% \author[1,+]{M. Elise Lauterbur}
% \author[3,*]{Ariella L. Gladstein}
 \author[4,*]{Graham Gower}
 \author[5*]{Nathaniel S. Pope}
 \author[5*]{Murillo F. Rodrigues}
\author[6]{Georgia Tsambos}
\author[34]{Linh N. Tran}
\author[2]{Maria Izabel A. Cavassim}
% \author[5,7]{Jeff Adrion}
% \author[5]{Saurabh Belsare}
% \author[8]{Arjun Biddanda}
% \author[5]{Victoria Caudill}
% \author[9]{Jean Cury}
% \author[10]{Ignacio Echevarria}
% \author[11]{Benjamin C. Haller}
% \author[12,13]{Ahmed R. Hasan}
\author[14,15]{Xin Huang}
% \author[16]{Leonardo Nicola Martin Iasi}
% \author[17]{Ekaterina Noskova}
% \author[18]{Jana Obšteter}
% \author[19]{Vitor Antonio Corrêa Pavinato}
% \author[20,21]{Alice Pearson}
% \author[22,23]{David Peede}
% \author[24]{Manolo F. Perez}
\author[5]{Chris C. R. Smith}
% \author[25]{Jeffrey P. Spence}
% \author[5]{Anastasia Teterina}
\author[5]{Silas Tittes}
\author[5]{Scott T. Small}
% \author[26]{Per Unneberg}
% \author[27]{Juan Manuel Vazquez}
% \author[28]{Ryan K. Waples}
% \author[29]{Anthony Wilder Wohns}
% \author[30]{Yan Wong}
% \author[31]{Franz Baumdicker}
% \author[32]{Reed A. Cartwright}
% \author[33]{Gregor Gorjanc}
\author[4]{Kirk E. Lohmueller}
\author[34]{Ryan N. Gutenkunst}
\author[30]{Jerome Kelleher}
\author[35]{Aaron P. Ragsdale}

\author[37]{Daniel R. Schrider}
\author[38]{Ilan Gronau}
\author[5,36]{Peter L. Ralph}
\author[5]{Andrew D. Kern}

@@

\affil[30]{\small{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery, University of Oxford, Oxford OX3 7LF, UK}}

@@

\affil[35]{\small{Department of Integrative Biology, University of Wisconsin-Madison, Madison WI, USA}}
\affil[36]{\small{Department of Mathematics, University of Oregon, Eugene OR 97402, USA}}
\affil[37]{\small{Department of Genetics, University of North Carolina at Chapel Hill, Chapel Hill NC 27599, USA}}
\affil[38]{\small{Efi Arazi School of Computer Science, Reichman University, Herzliya, Israel}}

@@

\section*{Introduction}
    \label{introduction}
    % natural selection
    Natural selection is a fundamental force in evolution, shaping the
    genetic diversity of populations and driving the adaptation of
    species to their environments. The effects of natural selection
    on genetic variation are complex, and can be difficult to disentangle
    from other evolutionary processes such as mutation, recombination,
    and genetic drift \cite[e.g.,][]{gillespie1991causes}.
    For instance changes in population size can lead to fluctuations
    in genetic diversity across a recombining chromosome 
    that can mimic the effects of selection \citep{simonsen1995properties, barton1998effect},
    % DRS -- I'm not sure Barton reference is right for this?
    and lead to spurious inferences about the strength and targets of genetic adaptation
    \cite{simonsen1995properties,akey2004population,nielsen2005genomic}.
    In turn selection can confound our ability to infer demographic 
    history from allele frequencies \citep{ewing2016consequences,schrider2016effects} and
    estimates of inverse coalescent rate \citep{schrider2016effects, johri2021impact, cousins2024accurate}.
    Thus it is imperative to joingly account for the effects of selection
    and demography when inferring evolutionary history from genetic data \citep{sheehan2016deep,johri2020toward}.
    However this is a challenging task from a modeling perspective \citep{johri2022prospect}.

    % simulation
    % DRS: lots more refs will be needed in this paragraph. added a few and am happy to help add more later if needed
    To meet the growing need for the interpretation,
    analysis, and exploration of realistic and complex evolutionary models
    the field of population genetics has increasingly turned to simulation.
    Simulation in population genetics has a long history 
    including both backward-in-time coalescent simulations
    \citep{kingman1982genealogy,hudson1983testing, hudson1990gene}
    and forward-in-time simulations of complex demography and selection
    \citep[e.g.,][]{gillespie1984molecular,thornton2014c++, haller2019slim}.
    The development of simulation tools has been driven by the need to
    understand the effects of complex evolutionary processes on genetic
    variation \citep[e.g.,][]{galloway2020few}, to provide a null model for hypothesis testing
    \citep[e.g.,][]{hudson1992statistical,hudson1994evidence,sabeti2002detecting}, to
    to explore the power and limitations of statistical methods \cite[e.g.,][]{przeworski2002signature},
    and increasingly to provide a basis for machine learning and other
    simulation-based inference methods \citep[e.g.,][]{pavlidis2010searching,lin2011distinguishing,kern2018diplos,mughal2019localizing,sanchez2021deep,wang2021automatic}.
    While this is so, joint simulation of complex demography and selection
    is challenging, and requires a deep understanding of the underlying
    evolutionary processes, as well as a number of parameter choices including
    the strength of selection, the distribution of fitness effects, and the
    recombination rate, as well as a parameterized model of demography.
    This is a daunting task for many researchers, and can be a barrier to
    the adoption of simulation-based methods in population genetics.
    Furthermore, sharing and comparing simulation results among studies can 
    itself be challenging, as different simulation tools may use different structures
    and parameter scalins for their models, and may not be directly comparable.
    This is a major limitation for the field, as it can make it difficult
    to assess the robustness of results, and to compare the results of
    different studies.

    % stdpopsim
    % DRS: lots more refs will be needed in this paragraph. added a few and am happy to help add more later if needed
    A lingering challenge with simulation in population genetics has been
    reproducibility and the ability to share and compare results among 
    researchers \citep[e.g.,][]{ragsdale2020lessons}.
    This challenge has been addressed in part by the development
    of community resources for sharing and distributing simulation software
    via the \stdpopsim project \citep{adrion2020community}. \stdpopsim
    provides a standardized interface for accessing a wide range of
    population genetic models, and has begun to be be widely adopted by the community. %maybe add citations here?
    While that is so, the original version of \stdpopsim did not include
    models of selection, which is a major limitation for more empirical
    applications of population genetic simulation. In particular, modeling selection
    through simulation is critical for understanding processes such
    as adaptation, the effects of selective sweeps \citep[e.g.][]{braverman1995hitchhiking,fay2000hitchhiking,przeworski2002signature,przeworski2005signature,schrider2015soft}, and the impact of
    background selection on genetic diversity
    \citep[e.g.][]{charlesworth1993effect,charlesworth1995pattern,williamson2002genealogy,ewing2016consequences,torres2020temporal}.
    Ideally, one would like
    to have a single, unified framework for simulating both neutral and
    non-neutral evolutionary processes, and to be able to compare the
    results of these simulations to empirical data in a manner that is
    both accessible to a wide range of researchers and highly reproducible. 
    Further the framework should include the complex realities of 
    genomes, including heterogenous recombination rates, 
    variation in the size and density of functional elements, and
    non-equilibrium demographic histories. Ideally, one should be able
    to model all of these features after estimates/annotations obtained
    from a species of interest \citep[e.g.][]{schrider2020background}.

    % goals
    In this manuscript we provide an overview of a major new addition
    to \stdpopsim---the inclusion of models of selection.
    We begin by describing the models of selection that we have implemented,
    as well as the parameter choices that are available to the user.
    This includes models of background selection, selective sweeps, and
    models of the distribution of fitness effects (DFE).
    Further we describe how these models can be combined with genomic
    annotations that are specific to a given species available
    through the \stdpopsim API, and combined with a parameterized model of
    demography and recombination to provide a realistic simulation of
    genetic variation in a population.
    We then provide a series of examples of how these models can be used
    to benchmark and compare the performance of different methods for
    inferring demographic history, the targets of recent positive selection,
    and the distribution of fitness effects from genetic data.

\section*{Implementing Selection in \stdpopsim}
    \label{selection}

    % implementing selection
    \stdpopsim previously provided a wide range of models of neutral
    demographic history from roughly two dozen species \citep{lauterbur2023expanding}.
    This was accomplished through the use of a standardized interface
    and data structure that allows for curated addition of new
    species and models, and can use either
    msprime \citep{Baumdicker2022} or SLiM \citep{haller2019slim}
    as the backend engine for efficient population genetic simulation.
    To extend \stdpopsim's functionality to include models of selection, 
    we introduce an important new class representing a specfic distribution of fitness effects
    to the \stdpopsim API: the \texttt{DFE} class. 
    This class provides an interface for specifying
    a distribution of the fitness effects of new mutations, and
    is added to a \texttt{Contig} object to provide a description
    of the fitness effects of mutations along a chromosome.
    A \texttt{DFE} object can apply to a specific interval set of a chromosome,
    and can be combined with other \texttt{DFE} objects to provide, 
    a rich model of how selection may differ along a chromosome. 
    Once specified a model of selection is then run in the same way
    but currently is limited to using SLiM simulation engine, which
    is perhaps the most flexible simulation engine for modeling selection currently available.

    As with other \stdpopsim models, when incorporating selection we aim to hew as closely as possible
    to realistic genomes, and provide public annotations of specices-specific functional
    elements that can be used along with a \texttt{DFE} object to
    readily provide realistic model of selection for a specific genome and set of
    populations. A schematic of the \stdpopsim catalog is shown in Figure \ref{fig:schematic}.
    From a users perspective, one can specify a species, a portion of the genome to simulate,
    a genetic map if available, a model of demography, and a model of selection that is
    composed of a set of annotations and a \texttt{DFE} object.

    % DRS: This figure could really use some Python API and comand line examples like in Figs 1B,C
    % form the original stdpopsim paper
    \begin{wrapfigure}[]{l}{0.4\textwidth}
        \vspace{-0.0cm}
        \includegraphics[width=\linewidth]{figures/schematics/catalog.pdf}
        \caption{\label{fig:schematic}
        A schematic of the \stdpopsim catalog for simulating genetic variation
        in a population. The user specifies a species, a portion of the genome to simulate,
        and optionally a genetic map, a model of demography, and a model of selection that 
        itself is composed of a distribution of fitness effects (DFE) and a set of functional
        annotations.}
    \end{wrapfigure}

    The vast majority of published DFE estimates are based on a limited number of species,
    so \stdpopsim currently has implemented published DFEs from
    four species: \textit{Arabidopsis thaleana} \citep{huber2018gene}, \textit{Drosophila melanogaster} \citep{ragsdale2016triallelic,huber2017determining},
    humans, and the vaquita porpoise \textit{Phocoena sinus} \citep{robinson2022critically}.
    While that is so, these DFEs can be mixed and matched with the other species
    in the catalog. For instance, one could simulate a butterfly species with a Gamma DFE originally
    estimated from humans, or a human population with a DFE estimated from \textit{Drosophila}.
    Furthermore the user can specify a custom DFE, and provide their own annotations
    of functional elements to simulate selection in a species for which we do not yet have 
    a published DFE included in the catalog. This flexibility allows for a wide range of
    models of selection to be simulated, and for the results of these simulations to be
    compared to empirical data or used as a null model for hypothesis testing.

    % sweep interface
    In addition to the \texttt{DFE} class, we have also implemented a class (\texttt{stdpopsim.ext.selective\_sweep})
    that enables selective sweeps to be simulated in \stdpopsim.
    This class augments a demographic model with an ``extended event''
    which conditions on the introduction of a selected mutation at a given time and position
    and with a given selection coefficient. Further the user can specify the minimum frequency
    of the selected allele at the time of sampling. As these extended events are implemented
    on top of the existing \stdpopsim API, they can be combined with other models of selection
    and demography to provide a rich model of the combined effects of multiple disparate evolutionary processes
    on genetic variation in a population.

    % current numbers of DFEs / species / etc
    At this release the \stdpopsim catalog comprises 24 species (i.e. genome representations)
    for which we have implemented XX demographic models, have XX genetic maps, and XX DFEs.
    Additions to the catalog are ongoing -- we point the reader to our previously published 
    report detailing this effort \citep{lauterbur2023expanding} and we welcome contributions from the
    community---we encourage those interested in contributing to visit https://popsim-consortium.github.io/stdpopsim-docs
    or contact one of the PopSim Consortium members for guidance.

\section*{Example Applications with Selection}
    \label{applications}
    In this section we provide a series of examples of how the new models of selection
    in \stdpopsim can be used to benchmark and compare the performance of different
    methods for population genetic inference.
    We focus on three main areas: demographic inference in the 
    single and multi-population settings, inference of the distribution of fitness effects,
    and the detection of selective sweeps, particularly in the context of background selection.

    \section*{Inference of $N_e(t)$ in the context of selection}
    One of the most common applications of population genetic inference is to estimate
    the effective population size over time, $N_e(t)$, from genetic data. This can be done
    using a variety of methods, including the pairwise sequentially Markovian coalescent
    (PSMC) \citep{li2011inference}, through use of the site frequency spectrum (SFS),
    as well as through identity by decent information. 
    While that is so, it has been well described that selection, because of its effects on
    local and more global patterns of genetic variation, can drive estimates of $N_e(t)$
    further from true census population sizes simply because selection increases the 
    rate of coalescence in the genome.

    Using \stdpopsim we can simulate genetic data under a model of selection and then 
    examine the performance of different methods for inferring $N_e(t)$ from these simulated 
    data. In Figure \ref{fig:1pop-human-demography} we show the results of such a benchmark
    where we have simulated genomes under a model of human out of Africa (OOA) demography
    with and without background selection on exons. For this simulation we have used
    a genetic map from the HapMap Project \citep{international2007second} along with a
    DFE for deleterious alleles (estimated by \cite{huber2017determining})
    that acts on exonic positions defined from the HAVANA group release 104 for the human genome. 
    These simulations were then used to estimate $N_e(t)$ using four methods: 1) \msmc \citep{Schiffels2020}, 
    2)\stairway method \citep{liu2020stairway}, 3) \gone \citep{santiago2020recent}, and 4) \smcpp \citep{terhorst2017robust}.
    In this and what follows we have performed 3 replicate whole genome simulations, so uncertainty in estimates might
    be roughly assessed, however we are mindful of wasted compute as thorough benchmarking
    isn't the main purpose of this manuscript.

    In Figure \ref{fig:1pop-human-demography} we show six panels, each of which shows the estimated $N_e(t)$
    from each independent sample in a five-population model of human out-of-Africa demography
    from Ragsdale \textit{et al.} (2019)
    with samples taken from three modern geographic groups (CEU, CHB, and YRI from left to right).
    The top row shows the results of simulations without selection, while the bottom row shows the results
    of simulations with a Gamma distributed DFE for the deleterious effects of mutations occurring exons. In each panel we show the true $N_e(t)$ in black, and the estimated $N_e(t)$
    in various colors for the four methods. Dotted lines on the bottom row are to visually reinforce that inference is done
    from a model with selection. We see that these methods generally perform reasonably well in the absence of selection, 
    although GONE seems to overestimate $N_e(t)$ in the more recent past for the CEU and CHB populations.
    Presumably this could be due to patterns of linkage disequilibrium that result from the combination
    of bottleneck and growth in these populations.
    In the presence of selection, different methods behave differently. 
    For instance stairwaiplot which uses the SFS to infer $N_e(t)$ seems to slightly underestimate $N_e(t)$
    but is reasonably close to the true $N_e(t)$ in the absence of selection.
    Likewise \smcpp seems to be more robust to the presence of selection,
    but there is some added noise to the returned estimates. 
    \msmc seems to be the most affected by the presence of selection, 
    estimates of $N_e(t)$ varying quite a bit from the true $N_e(t)$.
    % DRS: I don't agree with the interpretation in the previous two sentences (starting with "Likewise").
    % MSMC is already barfing a little bit even without selection, and it is not clear to me from these
    % plots that selection makes things any worse. Also, SMC++ is doing quite well without selection so maybe
    % the minor drop in accuracy is meaningful in relative terms.

    \begin{figure}[t]
        \centering
        \includegraphics[width=\textwidth]{figures/HomSap/OOA/estimated_Ne_t_final}
        \caption{
        \label{fig:1pop-human-demography}
        Performance of methods to infer $N_e(t)$ from a human out-of-Africa model \citep{ragsdale2019models}
        with and without background selection on exons. The top row shows estimates of $N_e(t)$ from simulations
        without selection, while the bottom row shows estimates of $N_e(t)$ from simulations with a Gamma distributed   
        DFE acting on exons. In each panel we show the true $N_e(t)$ in black, and the estimated $N_e(t)$ from four methods:    
        \msmc \citep{Schiffels2020}, \stairway \citep{liu2020stairway}, \gone \citep{santiago2020recent}, and \smcpp \citep{terhorst2017robust}.  
        }
    \end{figure}

    To highlight the ease of comparisons between different species and model enabled by \stdpopsim,
    we have also performed a similar benchmark using the vaquita porpoise \textit{Phocoena sinus}.
    In Figure \ref{fig:1pop-vaquita-demography} we show the results of $N_e(t)$ inference
    where we have simulated genomes under a model of vaquita demography from \textcite{robinson2022critically}.
    For these simulations we have used the genome structure and exon annotations from the vaquita genome assembly
    and a DFE from \textcite{robinson2022critically} that acts on exonic positions, however no genetic map 
    is currently available, so recombination rates are assumed to be constant across the genome.
    The demographic model is a simple two-epoch model where population size has declined in the recent past. 
    In this case we see that all methods perform well both in the absence and presence of selection,
    with a close match between the true $N_e(t)$ and the estimated $N_e(t)$ in all cases.

\begin{figure}[t]
    \centering
    \includegraphics[width=\textwidth]{figures/PhoSin/Vaquita2Epoch_1R22/estimated_Ne_t_final}
    \caption{
    \label{fig:1pop-vaquita-demography}
    Performance of methods to infer $N_e(t)$ from simulations of the vaquita porpoise genome under a single population
    model of declining population size \citep{robinson2022critically} with and without background selection on exons. 
    The top panel shows estimates of $N_e(t)$ from simulations
    without selection, while the bottom panel shows estimates of $N_e(t)$ from simulations with a Gamma distributed   
    DFE acting on exons. In each panel we show the true $N_e(t)$ in black, and the estimated $N_e(t)$ from four methods:    
    \msmc \citep{Schiffels2020}, \stairway \citep{liu2020stairway}, \gone \citep{santiago2020recent}, and \smcpp \citep{terhorst2017robust}.  
    }
\end{figure}

\section*{Estimation of the Distribution of Fitness effects}
    \label{dfe}
    Another common application of population genetic inference is to estimate the distribution of fitness effects (DFE)
    from genetic data. Here we show how the new models of selection in \stdpopsim can be used to benchmark and compare
    the performance of different methods for inferring the DFE from genetic data. Here we have performed simulations using the 
    human and vaquita porpoise models described above, with complex demographic histories and models of selection acting on exons.
    To remind the reader, the model of selection in each case is a Gamma distributed DFE for deleterious mutations.

    \begin{figure}[htbp]
        \centering
        \includegraphics[width=\linewidth]{figures/HomSap/OOA/HomSap_OutOfAfricaArchaicAdmixture_5R19_Gamma_K17_ensembl_havana_104_exons_DFE_plot}
        \caption{A comparison of methods for inferring the distribution of fitness effects (DFE) from genetic data.
        Simulations were performed using a human out-of-Africa model \citep{ragsdale2019models} with a Gamma distributed DFE
        acting on exons parameterized by a mean selection coefficient and a shape parameter. Estimates of the 
        parameters of the DFE are shown in the left ($\lvert E(s) \rvert $) and right hand panels (shape) respectively.
        The human model has three extant populations (CEU, CHB, and YRI), and parameter estimates from each
        sample are shown in different colors.
        Estimates from three different methods are shown: 1) \grapes \cite{galtier2016adaptive}, 2) \polydfe \citep{tataru2020polydfe},
        and 3) \dadi \citep{gutenkunst2009inferring}.}
        \label{fig:homsap-dfe.ooa}
    \end{figure}

    In Figures \ref{fig:homsap-dfe.constant} and \ref{fig:homsap-dfe.ooa} we show a comparison of estimates
    from three software packages for inferring the DFE from genetic data-- 
    \dadi \citep{gutenkunst2009inferring}, \polydfe \citep{tataru2020polydfe}, 
    and \grapes \citep{galtier2016adaptive}. 
    Figure \ref{fig:homsap-dfe.constant} shows estimates from a single, constant size population of humans, 
    while Figure \ref{fig:homsap-dfe.ooa} shows estimates from a model of human out-of-Africa demography.
    In a constant size population, all methods perform reasonably well, although \dadi seems to
    perform slightly worse than the other methods examined, at least for the small number of replicates we simulated here.
    Notably only \polydfe overlapped the true mean selection coefficient, whereas 
    the other methods slightly underestimated the strength of selection. 
    For the shape parameter in constant size simulations \grapes and \polydfe are roughly
    equivalent in their performance whereas \dadi seems to yield overestimates.

    In the out-of-Africa model (fig \ref{fig:homsap-dfe.ooa}) the performance of the methods is more mixed. 
    In particular variation in accuracy of the parameters estimated in each
    sample population depends on the methods used, and each method seems to 
    have its own biases with respect to the way in which population history 
    affects parameter estimates. For instance for \dadi we see that estimates of 
    the shape parameter are most accurate in the YRI sample, while for \polydfe
    estimates are consistently least accurate in the YRI sample. 
    With respect to the strength of selection, as in constant size populations,
    all methods are performing reasonably well, although \dadi seems to
    trail in terms of accuracy.

    %add PhoSin DFE figure here
    \begin{figure}
        \centering
        \includegraphics[width=0.8\textwidth]{figures/PhoSin/Vaquita2Epoch_1R22/PhoSin_Vaquita2Epoch_1R22_Gamma_R22_Phocoena_sinus.mPhoSin1.pri.110_exons_DFE_plot.pdf}
\includegraphics{figures/PhoSin/Vaquita2Epoch_1R22/hs_hist.pdf}
        \caption{
        \label{fig:vaquita-dfe}
        A comparison of methods for inferring the distribution of fitness effects (DFE) from genetic data.
        Simulations were performed using a model of vaquita porpoise demography \citep{robinson2022critically} with a Gamma distributed DFE
        acting on exons parameterized by a mean selection coefficient, a shape parameter, and a relationship between selection and domiance. Estimates of the 
        parameters of the DFE are shown in the left ($\lvert E(s) \rvert $) and right hand panels (shape) respectively.
        This vaquita model has a single extant population, labelled pop 0, and parameter estimates from each
        of the three different methods are shown: 1) \grapes \cite{galtier2016adaptive}, 2) \polydfe \citep{tataru2020polydfe},
        and 3) \dadi \citep{gutenkunst2009inferring,kim2017inference}.
        C) Shown is the simulated distribution of $2 h s$, where $h$ is the dominance coefficient, along with the simulated distribution of $s$ and the median inferred DFEs. The distribution of $2 h s$ is multimodal because the simulated relationship between $h$ and $s$ is not continuous.
        % RNG: For C, need to extract inferred values from pipeline.
        }
    \end{figure}

    In Figures \ref{fig:vaquita-dfe.constant} and \ref{fig:vaquita-dfe} we show a comparison of estimates
    from the same three software packages for inferring the DFE from genetic data, but for simulations of the
    vaquita porpoise genome. In the constant size model (fig \ref{fig:vaquita-dfe.constant}) we see that all methods
    perform uniformly worse that in the human genome simulations, which consistent underestimation of the mean selection coefficient    
    and overestimation of the shape parameter. 
A critical distinction is that the previous human simulations employed a DFE that assumed all selected mutations had additive effects (no dominance, $h = 1/2$), whereas the vaquita simulations use a DFE in which more deleterious mutations are more recessive ($h < 1/2$).
For all three software packages, the inferred mean and shape of the distribution of selection coefficients is consistent with the mean and shape of the distribution of $2 h s$, the product of the dominance coefficient and the selection coefficient (Fig.~\ref{fig:vaquita-dfe}C).
This is likely because deleterious alleles are typically at low frequency and thus heterozygous, where their selective effect is $h s$, and all these tools assume additivity  ($h = 1/2$).
Strongly deleterious alleles are often observed to be recessive in many species \citep{}, so these results suggest caution in interpreting DFE inferences.
%RNG: Could use Kirk's help with the best references here

\section*{Detection of Selective Sweeps}
    \label{sweeps}
    We next highlight how the new models of selection in \stdpopsim can be used to benchmark and compare
    the performance of different methods for detecting selective sweeps from genetic data. 
    Here we have performed simulations of human chromosome 1 under the three population out-of-Africa model from \cite{gutenkunst2009inferring}
    with purifying selection acting on exons under the Gamma-distributed DFE described above, the HapMap II genetic map \citep{international2007second},    
    but overlay on top of this a model of selective sweeps. In particular we have conditioned simulations on the introduction of a beneficial mutation
    at a given location (to be varied) in the genome at a given time, and with a moderately strong selection coefficient ($s = 0.03$; $2Ns \sim 600$).
    For each sweep location we performed 200 replicate simulations, sampling from them only if the selected allele reached a frequency of
    0.95 or greater. Rather than simulate the entire chromosome arm, replicate simulations were performed on 5cM regions of the chromosome
    at each of 100 different locations. As recombination rate varies across the genome, the size of the region in base pairs simulated varies,
    however the recombination distance, which is critical for the linked effects of a sweep, is held constant.
